Reinforcement Learning Produces Dominant Strategies for the Iterated Prisoner's Dilemma

### 2017-07-10 main.tex
Address @mojones comments.
Note I did not address:

>  page 3 "parameters of the neural network" should this be "weights of
the neural network" to match convention

I feel that parameters is fine but I'm easy.

The LookerUp strategy is based on a lookup tables and  encodes a set of
deterministic responses based on the opponent's first $n_1$ moves, the
opponent's last $m_1$ moves, and the players last $m_2$ moves. If $n_1 > 0$ then
the player has infinite memory depth, otherwise it has depth $\max{m_1, m_2}$.
This is illustrated diagrammatically in Figure~\ref{fig:lookerup}.

@@

This strategy type can be used to train arbitrary memory-$n$ strategies. A
memory one strategy called PSO Gambler Mem 1 was trained, with
probabilities $(\text{Pr}(\text{C}\;|\;\text{CC}),
                \text{Pr}(\text{C}\;|\;\text{CD}),
                \text{Pr}(\text{C}\;|\;\text{DC}),
                \text{Pr}(\text{C}\;|\;\text{DD})) = (1, 0.5217, 0, 0.121)$.
Though it performs well in standard tournaments (see
Table~\ref{tbl:standard_score})
it is not as good as the longer memory strategies, and is bested by a similar
strategy that also uses the first round of play: PSOGambler\_1\_1\_1.

@@

Training this strategy corresponds to finding mappings of states and histories
to an action and a state. Figure~\ref{fig:fsm_images} show two of the trained
finite state machines. The layout of state nodes is kept the same between
Figure~\ref{fig:fsm16} and~\ref{fig:fsm16noise} to highlight the effect of
different training environments. Note also that two of the 16 states are not
used, this is also an outcome of the training process.

\begin{figure}[!hbtp]
    \centering
    \begin{subfigure}[t]{.5\textwidth}
        \includestandalone[height=.3\textheight]{./assets/FSM16}
        \caption{Evolved\_FSM\_16: trained to maximise score in a standard
        tournament}
        \label{fig:fsm16}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.5\textwidth}
        \centering
        \includestandalone[height=.35\textheight]{./assets/FSM16noise}
        \caption{Evolved\_FSM\_16\_Noise\_05: trained to maximise score in a
        noisy tournament}
        \label{fig:fsm16noise}
    \end{subfigure}%
    \caption{Trained sixteen state finite state machine players}
    \label{fig:fsm_images}
\end{figure}

@@

A variant of finite state machine strategies are called hidden Markov models
(HMMs). Like the strategies based on finite state machines, these strategies
also encode an internal state however use probabilistic transitions based on the
prior round of play to other states and cooperate or defect with various
probabilities at each state. This is
shown diagrammatically in Figure~\ref{fig:hmm}. Training this strategy
corresponds to finding mappings of states and histories to probabilities of
cooperating as well as probabilities of the next internal state.

@@

    \caption{Comparison of collaboration rates for Standard Tournament Top 3
    (over 10000 repetitions)}
    \label{fig:comparison_cooperation_heatmaps_standard}
\end{figure}

@@

Finally, Table~\ref{tbl:noisy_ranks} and
Figure~\ref{fig:noisy_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. We see that the stochasticity
of the ranks understandably increases the DBS strategy never ranks lower than
second and wins 75\% of the time. The two strategies trained for noisy
tournaments rank in the top three 95\% of the time.
% TODO Possibly update this with more data.

@@

    \caption{Comparison of collaboration rates for the Noisy (5\%) Tournament Top 3
    (over 10000 standard tournaments)}
    \label{fig:comparison_cooperation_heatmaps_noisy}
\end{figure}

@@

The trained strategies (denoted by a \(^{*}\) in
Appendix~\ref{app:list_of_players} were trained using reinforcement
learning algorithms. The ideas of reinforcement learning can be attributed to
the original work of \cite{turing1950computing} in which the notion that
computers would learn by taking random actions but according to a distribution
that picked actions with high rewards more often. The two particular algorithms
used here:

>>>>

The LookerUp strategy is based on a lookup table and  encodes a set of
deterministic responses based on the opponent's first $n_1$ moves, the
opponent's last $m_1$ moves, and the players last $m_2$ moves. If $n_1 > 0$ then
the player has infinite memory depth, otherwise it has depth $\max(m_1, m_2)$.
This is illustrated diagrammatically in Figure~\ref{fig:lookerup}.

@@

This strategy type can be used to train arbitrary memory-$n$ strategies. A
memory one strategy called PSOGamblerMem1 was trained, with
probabilities $(\text{Pr}(\text{C}\;|\;\text{CC}),
                \text{Pr}(\text{C}\;|\;\text{CD}),
                \text{Pr}(\text{C}\;|\;\text{DC}),
                \text{Pr}(\text{C}\;|\;\text{DD})) = (1, 0.5217, 0, 0.121)$.
Though it performs well in standard tournaments (see
Table~\ref{tbl:standard_score})
it is not as good as the longer memory strategies, and is bested by a similar
strategy that also uses the first round of play: PSOGambler\_1\_1\_1.

@@

Training this strategy corresponds to finding mappings of states and histories
to an action and a state. Figure~\ref{fig:fsm_images} shows two of the trained
finite state machines. The layout of state nodes is kept the same between
Figure~\ref{fig:fsm16} and~\ref{fig:fsm16noise} to highlight the effect of
different training environments. Note also that two of the 16 states are not
used, this is also an outcome of the training process.

\begin{figure}[!hbtp]
    \centering
    \begin{subfigure}[t]{.5\textwidth}
        \includegraphics[height=.3\textheight]{./assets/FSM16.pdf}
        \caption{Evolved\_FSM\_16: trained to maximise score in a standard
        tournament}
        \label{fig:fsm16}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{.5\textwidth}
        \centering
        \includegraphics[height=.35\textheight]{./assets/FSM16noise.pdf}
        \caption{Evolved\_FSM\_16\_Noise\_05: trained to maximise score in a
        noisy tournament}
        \label{fig:fsm16noise}
    \end{subfigure}%
    \caption{Trained sixteen state finite state machine players}
    \label{fig:fsm_images}
\end{figure}

@@

A variant of finite state machine strategies are called hidden Markov models
(HMMs). Like the strategies based on finite state machines, these strategies
also encode an internal state. However, they use probabilistic transitions based on the
prior round of play to other states and cooperate or defect with various
probabilities at each state. This is
shown diagrammatically in Figure~\ref{fig:hmm}. Training this strategy
corresponds to finding mappings of states and histories to probabilities of
cooperating as well as probabilities of the next internal state.

@@

    \caption{Comparison of cooperation rates for Standard Tournament Top 3
    (over 10000 repetitions)}
    \label{fig:comparison_cooperation_heatmaps_standard}
\end{figure}

@@

Finally, Table~\ref{tbl:noisy_ranks} and
Figure~\ref{fig:noisy_ranks_boxplot} show the ranks (based on median score)
of each strategy over the repeated tournaments. We see that the stochasticity
of the ranks understandably increases relative to the Standard tournament. An
exception is the top three strategies, for example, the DBS strategy never ranks
lower than
second and wins 75\% of the time. The two strategies trained for noisy
tournaments rank in the top three 95\% of the time.
% TODO Possibly update this with more data.

@@

    \caption{Comparison of cooperation rates for the Noisy (5\%) Tournament Top 3
    (over 10000 standard tournaments)}
    \label{fig:comparison_cooperation_heatmaps_noisy}
\end{figure}

@@

The trained strategies (denoted by a \(^{*}\) in
Appendix~\ref{app:list_of_players}) were trained using reinforcement
learning algorithms. The ideas of reinforcement learning can be attributed to
the original work of \cite{turing1950computing} in which the notion that
computers would learn by taking random actions but according to a distribution
that picked actions with high rewards more often. The two particular algorithms
used here:
